Causal inference and the detection of coevolutionary contingencies using dynamic phylogenetic models

### 2024-08-30 manuscript.Rmd
Update prior for simulations

---
title: "Causal inference and the detection of coevolutionary contingencies using dynamic phylogenetic models"
author: "E. Ringen, J. S. Martin, S. Claessens, & A. V. Jaeggi"
date: "`r Sys.Date()`"
output: 
  pdf_document:
    number_sections: true
bibliography: references.bib
link-citations: yes
linkcolor: blue
urlcolor: black
figsintext : yes 
header-includes:
  \usepackage{fixltx2e}
  \usepackage{hyperref}
  \newcommand\figref[1]{{\hypersetup{linkcolor=blue}\autoref{#1}}}
---

\begin{center}
Abstract text here...
\end{center}

@@

Phylogenetic comparative methods (PCMs) are commonly used to study the
co-evolution of organismal traits, spanning topics such as anatomy and 
physiology [@Dunn2015; @Garland2005; @Navalon2019; @OConnor2022; @Thayer2018], 
life history and behavior [@Bielby2007; @Clayton1994; @MacLean2012; 
@Salguero2016], and cultural evolution [@Mace2005; @Navarrete2016; @Watts2016]. 
"Coevolution" in its broadest sense refers to repeatable patterns of trait 
covariation over time, which can be investigated using a diverse family of 
statistical techniques depending on the research question and type of data 
available [@Garamszegi2014; @Harvey1991; @Nunn2011]. Many PCMs focus on 
understanding how traits coevolve by estimating evolutionary correlations and 
rates of change across phylogenetic trees, which can be used to better 
understand natural history and inform ancestral state reconstructions. However,
evolutionary correlation does not imply evolutionary causation between traits 
[@Shipley2016]. Thus, PCMs facilitating inference of why traits coevolve and 
change together across time are indispensable tools for testing 
macroevolutionary and ecological theory.

@@

We begin by briefly reviewing the benefits and constraints of current PCMs, 
particularly with regard to causal inference, exploring how their limitations 
motivate further methods development. We then formally and conceptually 
introduce a novel class of dynamic phylogenetic models. We provide a worked 
synthetic example of coevolution between two continuous traits -- illustrated as
female promiscuity and sperm length [@Fitzpatrick2009] -- to assess the accuracy
and uncertainty of inferences made with dynamic phylogenetic models across a 
reasonable range of sample sizes for phylogenetic research. We also provide an 
accompanying coding tutorial using our **coevolve** R package to aid empiricists
in easily applying basic dynamic phylogenetic models to their own datasets. We 
then demonstrate the generality and flexibility of our method with two empirical
applications, which extend the model to complex, high-dimensional scenarios 
using dynamic latent variables. In particular, we show how our method can 
generate insights into the causal dynamics of trait coevolution across both 
genetic and cultural evolutionary timescales, using (1) a comparative dataset on
primate brain size, sociality, diet, and life history traits [@DeCasien2017], 
which reduced to two coevolving latent variables, and (2) two studies on the 
evolution of social complexity across a global and a regional sample of 
pre-industrial human societies [@Ringen_preprint; @Sheehan2023].

@@

Fundamental to PCMs is the adjustment of raw trait associations for shared 
evolutionary history using a phylogenetic tree (or set of trees) and a 
statistical model. In a basic sense, phylogenetic adjustment is crucial for 
facilitating causal inference, as shared evolutionary history tends to generate 
trait correlations among closely related species with similar phenotypes, 
creating the illusion of convergent co-evolution even when traits evolve 
independently. Adjustment for phylogeny reduces bias and variance due to 
clustering over the tree, thus reducing the risks of type I (false-positive), 
type II (false-negative), type M (magnitude), and type S (sign) errors during 
statistical inference (*CITE*). Nonetheless, adjustment for phylogeny is not a 
magic fix for all sources of unobserved confounding, nor does it guarantee that 
resulting estimates are causally interpretable (*CITE*). Therefore, while all 
PCMs employ statistical techniques that reduce bias, these methods vary widely 
in the degree to which they isolate the causal pathways by which traits coevolve 
over time, with most commonly used methods focusing largely on evolutionary 
correlation (\figref{fig:dags}a) rather than causation (\figref{fig:dags}b-e).

```{=latex}
\begin{figure}
```

```{r dags, echo=FALSE, out.width='100%'}
knitr::include_graphics("figures/dags.png")
```

```{=latex}
\caption{Explanatory models of trait coevolution. \label{fig:dags}}
\textit{Note:} Examples of distinct formal approaches to describing and 
explaining patterns of coevolution among traits (bold letters), with important 
properties represented in each graph highlighted by blue arrows and text. 
Approaches range in complexity from (a) simple models of phylogenetic 
correlations (indicated by bidirectional arrows), which are useful for 
predictive purposes but fail to explain the causes of trait associations; to 
(b-e) directed acyclic graph models, which can be used to directly test the 
causal effects (directed arrows) driving trait associations across a phylogeny. 
As explained in Box 2 below, explicit causal models are crucial for deciding 
which traits should be included or excluded from a multivariate analysis to 
avoid potential biases due to phenomena such as so-called forks (b) and 
colliders (c). Causal models can also be further distinguished by whether they 
model relationships among traits as static (b-c) or dynamic (d-e) effects. Only 
dynamic models can be used to account for feedback processes 
(t\textsubscript{-1} → t → t\textsubscript{+1}) generated by reciprocal 
causation (blue arrows) among traits and autoregressive effects within traits 
(grey arrows) over time (d). For high dimensional problems, it is important to 
also consider whether inclusion of latent causes into the dynamic DAG (e), 
capturing dimensions of evolutionary integration among multiple traits, can 
provide a more parsimonious and theoretically insightful causal model (see 
\autoref{sec:empirical}).
\end{figure}
```

@@

@@

@@

@@

@@

@@

@@

## Empirical applications

\label{sec:empirical}

### Brain size evolution in primates

### Evolution of social complexity across human societies

>>>>

---
title: "Causal inference and the detection of coevolutionary contingencies using dynamic phylogenetic models"
author: "E. Ringen, J. S. Martin, S. Claessens, & A. V. Jaeggi"
date: "`r Sys.Date()`"
output: 
  pdf_document:
    number_sections: true
    includes:
      in_header: header.tex
bibliography: references.bib
link-citations: yes
linkcolor: blue
urlcolor: blue
figsintext : yes
---

\begin{center}
Phylogenetic comparative methods are widely used to study the coevolution of
biological and cultural traits over evolutionary time. Among these methods, 
some of the most popular are phylogenetic generalised linear mixed models, 
phylogenetic path analysis, and Pagel's 'discrete' method. While these methods
have clear benefits under specific scenarios, they also have important
limitations, such as unrealistic assumptions about the evolutionary process,
a focus on evolutionary correlation rather than causation, and limits on the 
number and type of traits that researchers can include. Here, we develop a
novel class of Bayesian *dynamic phylogenetic models* (DPMs) that overcomes
these limitations and synthesizes the strengths of existing phylogenetic 
comparative methods into a single generalizable framework for dynamic
phylogenetic causal inference. We formally and conceptually introduce this
class of model, which is applicable to any number of traits following any
response distribution. We provide a short tutorial for fitting these models
with the `coevolve` R package, a user-friendly package purpose-built for 
fitting DPMs in Stan. We highlight previous studies that have used DPMs to 
make inferences about trait coevolution, including analyses of brain size 
in non-human primates and social complexity in human societies, and suggest
some directions for future model development and research.
\end{center}

@@

Phylogenetic comparative methods (PCMs) are commonly used to study the
coevolution of organismal traits, spanning topics such as anatomy and 
physiology [@Dunn2015; @Garland2005; @Navalon2019; @OConnor2022; @Thayer2018], 
life history and behavior [@Bielby2007; @Clayton1994; @MacLean2012; 
@Salguero2016], and cultural evolution [@Mace2005; @Navarrete2016; @Watts2016]. 
"Coevolution" in its broadest sense refers to repeatable patterns of trait 
covariation over time, which can be investigated using a diverse family of 
statistical techniques depending on the research question and type of data 
available [@Garamszegi2014; @Harvey1991; @Nunn2011]. Many PCMs focus on 
understanding *how* traits coevolve by estimating evolutionary correlations and 
rates of change across phylogenetic trees, which can be used to better 
understand natural history and inform ancestral state reconstructions. However,
evolutionary correlation does not imply evolutionary causation between traits 
[@Shipley2016]. Thus, PCMs facilitating inference of *why* traits coevolve and 
change together across time are indispensable tools for testing 
macroevolutionary and ecological theory.

@@

We begin by briefly reviewing the benefits and constraints of current PCMs, 
particularly with regard to causal inference, exploring how their limitations 
motivate further methods development. We then formally and conceptually 
introduce a novel class of *dynamic phylogenetic models* (DPMs). We provide a 
worked synthetic example of coevolution between two continuous traits -- 
illustrated as female promiscuity and sperm length [@Fitzpatrick2009] -- to 
assess the accuracy and uncertainty of inferences made with dynamic phylogenetic
models across a reasonable range of sample sizes for phylogenetic research. We 
also provide an accompanying coding tutorial using our `coevolve` R package to
aid empiricists in easily applying basic dynamic phylogenetic models to their 
own datasets. We then demonstrate the generality and flexibility of our method 
with two empirical applications, which extend the model to complex, 
high-dimensional scenarios using dynamic latent variables. In particular, we 
show how our method can generate insights into the causal dynamics of trait 
coevolution across both genetic and cultural evolutionary timescales, using (1) 
a comparative dataset on primate brain size, sociality, diet, and life history 
traits [@DeCasien2017], which reduced to two coevolving latent variables, and 
(2) two studies on the evolution of social complexity across a global and a 
regional sample of pre-industrial human societies [@Ringen_preprint; 
@Sheehan2023].

@@

Fundamental to PCMs is the adjustment of raw trait associations for shared 
evolutionary history using a phylogenetic tree (or set of trees) and a 
statistical model. In a basic sense, phylogenetic adjustment is crucial for 
facilitating causal inference, as shared evolutionary history tends to generate 
trait correlations among closely related species with similar phenotypes, 
creating the illusion of convergent coevolution even when traits evolve 
independently. Adjustment for phylogeny reduces bias and variance due to 
clustering over the tree, thus reducing the risks of type I (false-positive), 
type II (false-negative), type M (magnitude), and type S (sign) errors during 
statistical inference (*CITE*). Nonetheless, adjustment for phylogeny is not a 
magic fix for all sources of unobserved confounding, nor does it guarantee that 
resulting estimates are causally interpretable (*CITE*). Therefore, while all 
PCMs employ statistical techniques that reduce bias, these methods vary widely 
in the degree to which they isolate the causal pathways by which traits coevolve 
over time, with most commonly used methods focusing largely on evolutionary 
correlation (\autoref{fig:dags}a) rather than causation (\autoref{fig:dags}b-e).

```{r dags, echo=FALSE, fig.align='center', fig.cap="\\textit{Explanatory models of trait coevolution.} Examples of distinct formal approaches to describing and explaining patterns of coevolution among traits (bold letters), with important properties represented in each graph highlighted by blue arrows and text. Approaches range in complexity from (a) simple models of phylogenetic correlations (indicated by bidirectional arrows), which are useful for predictive purposes but fail to explain the causes of trait associations; to (b-e) directed acyclic graph models, which can be used to directly test the causal effects (directed arrows) driving trait associations across a phylogeny. Explicit causal models are crucial for deciding which traits should be included or excluded from a multivariate analysis to avoid potential biases due to phenomena such as so-called forks (b) and colliders (c). Causal models can also be further distinguished by whether they model relationships among traits as static (b-c) or dynamic (d-e) effects. Only dynamic models can be used to account for feedback processes (t\\textsubscript{-1} → t → t\\textsubscript{+1}) generated by reciprocal causation (blue arrows) among traits and autoregressive effects within traits (grey arrows) over time (d). For high dimensional problems, it is important to also consider whether inclusion of latent causes into the dynamic DAG (e), capturing dimensions of evolutionary integration among multiple traits, can provide a more parsimonious and theoretically insightful causal model (see \\autoref{sec:empirical}).", out.width='100%'}
knitr::include_graphics("figures/dags/dags.png")
```

@@

Phylogenetic generalized linear (mixed) models are linear regression models used
to quantify trait covariance between species or populations due to shared 
evolutionary history (likewise for their predecessor, independent contrasts) 
[@Blomberg2012; @Grafen1989; @Hadfield2010; @Lynch1991; @Symonds2014]. The most 
commonly assumed model of trait coevolution in these regression models is 
Brownian motion (also called a Wiener process, *CITE*), which assumes that the 
trait variance for a given species or population is proportional to the total 
branch length of the tree leading from root to tip, and the covariance between 
any two species is proportional to their amount of shared history. This process
is often biologically interpreted as reflecting patterns of evolutionary drift 
(or neutral evolution) across a phylogenetic tree. Empirically, the actual 
degree of covariance between related species or populations may be less than
expected under a pure Brownian motion model, such as when evolution is rapid or,
more mundanely, when measurement error is substantial. In these cases, the 
"phylogenetic signal" will be weak such that even closely related taxa may have 
dramatically different trait values [@Blomberg2003; @Kamilar2013], motivating 
various branch length transformations to the basic model to account for this 
greater independence among taxa [@Pagel1999]. In cases where traits do not 
evolve according to a modified Brownian motion process, other adjustments can 
also be made to the basic phylogenetic generalized linear model to modify the 
expected rate change in trait (co)variation. For instance, early burst models 
of adaptive radiation allow the rate of change to decrease over time (*CITE*), 
while Ornstein-Uhlenbeck models characterize evolutionary change as the product
of both stochastic and deterministic processes (*CITE*).

Despite their differences, attempts to use parameters from any of these models
for causal inference share often unstated and undesirable assumptions, namely 
that there is no selection acting on the traits included as predictors, that all 
predictor traits are independent and direct causes of the response trait, and 
that there is no reciprocal causation between traits over time (see 
\autoref{fig:dags}). These are strong assumptions that are likely to be violated
in most empirical datasets, limiting the applicability of phylogenetic 
generalized linear (mixed) models for theory testing and development. In 
addition, the most commonly used implementation of these models, phylogenetic
generalized least squares regression, suffers from various other constraints 
such as being limited to traits with Gaussian errors, overfitting due to a lack
of parameter regularization, and a failure to accommodate many common data 
features such as repeated measures of species trait values, missing data, and 
measurement error. While multilevel/mixed-effects models can be used to address 
these statistical concerns [@Hadfield2010; @Ives2010; @Ives2011; @Lynch1991; 
@Martin2020; @Ringen2019], they nonetheless share the same basic limitations 
with regard to causal inference and are arguably not ideal for modelling 
adaptation [@Hansen2014] (Box \ref{box:evolutionary-regression-coefficients}).

\mybox[label=box:evolutionary-regression-coefficients]{Evolutionary regression coefficients underestimate causal effects}
An unintuitive aspect of traditional PCMs is that regression coefficients 
estimated using cross-sections of contemporary species are used to infer the 
change in some trait $Y$ in response to another trait $X$ over evolutionary 
time. The slope of a response trait on a predictor trait is known as the 
"evolutionary regression coefficient" [@Pagel1993]. For the evolutionary
regression coefficient to be interpreted as a causal effect, we must, as with 
all statistical models, assume that there are no unmeasured confounders and that
our parametric model is correct. But there is an additional caveat: because 
selection is (generally) a gradual process, the total causal effect of one trait
on another -- which we define as the change in the optimal trait value of $Y$ as
a function of the value of $X$ [see @Scholkopf2013 for a closely related 
definition of causal effects in systems of ordinary differential equations] -- 
can take a long time to be fully realized. As such, we capture species in the 
process of adaptation rather than at equilibrium. Consequently, traditional 
methods underestimate the total causal effect of one trait another. This 
attenuation bias is a joint product of the strength of selection on the response
trait and the rate of change in the predictor trait. Intuitively, if one trait 
changes too quickly, then the other will always be playing catch-up. See 
\autoref{fig:evo-reg-coef} for an illustration of this process. A solution to 
this problem is to move beyond traditional regression-based PCMs and explicitly
model trait change and the adaptive process using dynamic phylogenetic models.
\endmybox

```{r evo-reg-coef, echo=FALSE, fig.align='center', fig.cap='\\textit{Gradual adaptation flattens the evolutionary regression coefficient.} Ancestral species ("A") evolve towards their optimal trait Y value, which depends on the value of trait X shown on the solid line. Ancestors speciate and their descendants ("D") inherit their maladapted trait values, resulting in a slope between the two traits (dashed line) that is flatter than the true relationship. Redrawn from Figure 14.4 in Hansen (\\protect\\hyperlink{ref-Hansen2014}{2014}).', out.width='60%'}
knitr::include_graphics("figures/evolutionary_regression_coefficient.png")
```

@@

Fortunately, a major innovation in statistical research in recent decades has 
been to realize that even conventional regressions with observational data can 
be used to infer and test for the presence of causal relationships once 
appropriate assumptions are encoded into the model structure [@Shipley2016]. 
This is because specific causal models, usually represented as directed acyclic
graphs (DAGs, see \autoref{fig:dags}, Box 
\ref{box:evolutionary-regression-coefficients}), imply specific patterns of 
correlation (or covariance) among variables [@Pearl2009; @Rubin2005]. An 
elaborate toolbox now exists to determine the correct test of causal hypotheses
given a DAG. This has become the state-of-the-art in fields like epidemiology 
and is increasingly taught and used in evolutionary biology and ecology 
[@Deffner2022; @McElreath2020; @Shipley2016; @Warrell2020].

If one wants to test for a causal effect of $X$ (say diet) on $Y$ (say brain 
size), one has to carefully consider the relationship of potential third 
variables $Z$ (say sociality or life history), which, depending on the structure
of the DAG, should either be included in the regression analysis or not (see 
Figure 5). These considerations are what distinguish causal inference from 
standard predictive applications of multivariate regression. For instance, most 
researchers know that if $Z$ causes both $X$ and $Y$ (a "fork", 
\autoref{fig:dags}b), $Z$ should be included to remove spurious associations 
between $X$ and $Y$. However, if $Z$ is a collider (both $X$ and $Y$ cause $Z$; 
\autoref{fig:dags}c), $Z$ should not be included in the regression model, as 
doing so introduces a spurious association between $X$ and $Y$. For example, if
diet ($X$) and brain size ($Y$) independently caused longevity ($Z$) but were 
otherwise uncorrelated, conditioning on longevity would introduce a spurious 
negative association between diet and brain size. This further implies that the
appropriate set of covariates for testing a causal effect of $X$ on $Y$ may not
be the same as the set for testing the causal effect of $Z$ on $Y$. 
Nevertheless, all coefficients in a multivariate regression are commonly
interpreted as causal effects, a phenomenon known as the "table 2 fallacy" 
[@Westreich2013]. Most comparative analyses have not yet used the causal 
inference toolbox and typically include several variables in a phylogenetic 
generalized least squares model without formal causal justification. This 
implicitly assumes that all variables are independent causes, which is often 
problematic.

Phylogenetic path analysis represents a particular application of the causal 
inference toolbox that overcomes some of the issues outlined above for 
phylogenetic generalized least squares [@GonzalezVoyer2014; @vonHardenberg2013].
With phylogenetic path analysis, researchers pre-define specific causal models,
ideally avoiding colliders and other traps (\autoref{fig:dags}), and compare 
their fit to the data. Specifically, a DAG implies a set of variables that 
should be statistically independent, which is then tested using phylogenetic 
regression, such that several causal hypotheses can be compared. For instance, 
@Navarrete2016 presented several phylogenetic path analysis models that all 
included links between life history and social group size with brain size, with
variable direct or indirect effects of diet and technical intelligence. Thus, 
phylogenetic path analysis is well-suited for comparing existing hypotheses, 
with DAGs generated "top-down" from prior theory or empirical literature
[@GonzalezVoyer2014].

However, the applicability of phylogenetic path analysis is limited to static 
DAGs lacking any reciprocal effects over time, such as \autoref{fig:dags}b and c 
where $X$, $Y$, and $Z$ do not have any temporal structure that could be used
to resolve reciprocal causal effects. Such models assume a constant direction of
effect between two variables, with bidirectional causal effects only being able
to be described as unresolved patterns of correlation (\autoref{fig:dags}a). 
This is an issue because positive (or negative) evolutionary feedback loops are
predicted by many theoretical models in response to life history tradeoffs and 
various forms of frequency- and density-dependence, suggesting that reciprocal
causation is likely to be a common phenomenon explaining patterns of trait 
coevolution. For example, @McNamara2022 presents a model of the coevolution of
paternal care and extrapair paternity, which is driven by a frequency-dependent
feedback loop over evolutionary time between males' tendency to seek extrapair
copulations and the benefits of their paternal care. The inability to directly
test for such dynamic feedback effects is a fundamental limitation of 
phylogenetic path analysis and any other phylogenetic method reliant on static 
causal models, thus motivating the development of dynamic methods.

@@

To date, Pagel's [-@Pagel1994] "discrete" method is the only general application
of dynamic causal inference to phylogenetic data. By reconstructing the 
coevolutionary sequence of two traits along the branches of a phylogeny, 
researchers can infer causality and potentially reciprocal causality through 
temporal contingency, referred to as "Granger causality" in economics 
[@Granger1969]. For example, researchers can infer that $X$ evolved first and 
then made the evolution of $Y$ more likely. This method has resulted in many 
high-profile publications [@Cornwallis2017; @Fitzpatrick2009; @Kappeler2019; 
@Sheehan2018; @Shultz2011; @Watts2016], all of which are built on the stronger 
causal inference of temporal contingency.

Unfortunately, despite its innovative approach to causal inference, Pagel's 
method is fundamentally limited to investigating the coevolution of binary 
traits, where dynamic evolutionary transitions can be conceptualized in terms of
probabilities of switching between two discrete states. Application of this 
method to evolutionary topics involving continuous phenomena (e.g. morphology,
life history, brain size, etc.) thus requires that researchers falsely 
dichotomize naturally occurring variation in traits [e.g. @Cornwallis2017; 
@Fitzpatrick2009; @Sheehan2018; @Watts2016], which is well-established as an 
undesirable statistical practice leading to loss of power, biased effect sizes,
and inconsistent results [@Dawson2012; @Royston2006]. For instance, 
@Fitzpatrick2009 dichotomized continuously-measured sperm length and speed by 
classifying traits below the species mean as "low" and those above the species 
mean as "high", and they collapsed a four-point scale of female promiscuity, 
based on behavioral and paternity data, into "low" (levels 1 and 2) and "high" 
(levels 3 and 4). Using Pagel's method, they then showed that sperm evolved to 
be faster and longer in response to increases in female promiscuity.

The limitation of Pagel's method to binary traits is further exacerbated when 
studying the (co)evolution of multiple continuous traits, which often motivates
the incorporation of latent variables into a dynamic causal analysis 
(\autoref{fig:dags}e). Latent variables are widely used in biology to capture
theoretically pertinent constructs that are difficult to directly operationalize
and quantify using a single measurement, such as size and shape dimensions in 
morphometrics (*CITE*), environmental quality and climate metrics in ecology
(*CITE*), life history variation [@Stott2024], and canonical axes of 
correlational selection in quantitative genetics (*CITE*). While latent 
variables are sometimes used pragmatically as placeholders for deeper causal 
analysis [@Shipley2016], many evolutionary models directly conceptualize and 
model causal effects occurring among latent variables. For instance, 
life-history traits are theorized to evolve along a latent fast-slow continuum 
[@Bielby2007; @Healy2019; @Stott2024; @Wright2019] that can both cause and in 
turn be caused by the correlated evolution of a niche complexity continuum 
reflecting differences in brain size and cognitive ability, social group size, 
and foraging behavior [@Dunbar2017; @Miller2019; @Navarrete2016] (see 
\autoref{sec:brain-size-evo}). While prior research has used methods such as 
phylogenetic principal component analysis (pPCA; *CITE*) to apply Pagel's method
to latent variables [e.g. @Cornwallis2017], this approach still requires falsely
dichotomizing latent variables and thus suffers from the same statistical biases
as with any other continuous measure. Synthesizing continuous and multivariate
methods with dynamic causal inference thus remains a major challenge for 
contemporary PCMs.

@@

The limitations of current PCMs can be overcome by synthesizing their respective
strengths into a single generalizable framework for dynamic phylogenetic causal
inference, applicable to both discrete and continuous traits. We have attempted
to do so through the development of a novel class of PCMs we refer to simply as
dynamic phylogenetic models (DPMs) of trait coevolution. Our DPM is driven by a 
multivariate stochastic differential equation, similar to a multivariate 
Ornstein-Uhlenbeck (OU) process (see Box \autoref{box:ou-model} for an overview
of the basic OU model) with greater model flexibility and complexity than is 
permitted by standard implementations. Our approach arose from two complementary
goals: first, to extend previous Bayesian implementations of the multivariate OU
model such that the optimal trait value $\theta$ is updated dynamically (rather
than assuming piecewise constant, see Box \autoref{box:ou-model}). Second, we 
wanted a general form for the optimal trait value $\theta$ that could be used to
assess directionality (X → Y vs. Y → X) and contingencies (X, then Y) in 
evolution, akin to Pagel's [-@Pagel1994] method but without restrictions on the
type or number of traits.

\mybox[label=box:ou-model]{The Ornstein-Uhlenbeck (OU) model}
The OU model is a mean-reverting, stationary Gauss-Markov process. It describes
change in a trait due to both Gaussian noise and reversion towards some central
value. In some fields, such as economics, the mean reversion component is called
"drift" and the stochastic noise is called "diffusion" (i.e., the Vasicek model,
*CITE*). In evolutionary biology, the mean-reverting quality is loosely 
interpreted as "selection" and accordingly the Gaussian noise is labeled "drift"
(*CITE*). The basic OU form is

$$
dy(t) = \alpha(\theta - y_{t}) + \sigma dW(t)
$$

where $\alpha$ controls the strength of mean-reverting selection, $\theta$ is 
the mean trait value, and $\sigma$ controls the strength of drift. When 
$\alpha = 0$, the model is pure drift Brownian motion. The simplest OU models 
assume a single evolutionary optimum ($\theta$), or estimate an ancestral 
optimum along with a global optimum. More elaborate OU models imagine that 
$\theta$ changes as a function of other variables, turning it into a 
coevolutionary process with varying selection regimes (i.e., the Hansen model; 
*CITE*). These approaches exploit the fact that, if selection regimes are 
piecewise-constant (the optimum is the same within each segment of the 
phylogenetic tree but allowed to vary at branching points), the OU process can 
be discretized, giving a closed form solution for the expectation and covariance
matrix of a trait. Butler and King (*CITE*) provided a maximum-likelihood 
algorithm for this approach, and a Bayesian implementation was developed by 
Ross et al. (*CITE*). Depending on the research question, the piecewise constant 
assumption can be quite restrictive and a poor approximation when tree segments
are long, due to the assumption of a constant optimum within each discrete tree
segment.

\endmybox

@@

Rather than estimating a phylogenetic variance-covariance matrix as is standard
for PCMs, we instead partition the deterministic and stochastic components of 
the coevolutionary process, adapting the continuous-time structural equation 
modeling approach of @Driver2017 to the phylogenetic context. The evolutionary 
history of any species or population is modeled as a time series where the 
deterministic dynamics of an OU process play out over the length of each tree 
segment, and the stochastic drift components (which are by definition orthogonal
to selection) are added to the end of each segment as independent samples from 
the standard normal distribution, scaled by the expected covariance for a given 
segment duration $t(s)$

\begin{equation}
  \label{eq:ou-evo-model}
  d\eta(t) = (\textbf{A}\eta(t) + \textbf{b}) + \textbf{G}dW(t)
\end{equation}

Here $\eta(t)$ is a vector of the latent variables at time $t$, the matrix 
$\textbf{A}$ represents "selection" with autoregressive terms on the diagonal
equivalent to $\alpha$ in the univariate OU process (Box 
\autoref{box:ou-model}) and off-diagonals representing the effect of each trait
on the others (e.g., $\textbf{A}[2,1]$ represents the effect of $\eta_1$ on 
$\eta_2$), and $\textbf{b}$ is a vector of continuous time intercepts that, 
along with $\textbf{A}$, determine the asymptotic values of $\eta$. The matrix 
$\textbf{G}$ is the Cholesky decomposition of the "drift" covariance matrix 
$\textbf{Q}$, such that $\textbf{Q} = \textbf{G}\textbf{G}'$, which scales the
stochastic Weiner process. The square root of the diagonals in $\textbf{Q}$ are
equivalent to $\sigma$ in the OU process (Box \autoref{box:ou-model}). Although
in other types of time-series analyses it is possible to estimate the 
off-diagonals of $\textbf{Q}$ (i.e., the covariance of the stochastic drift 
terms), it is not possible to simultaneously estimate them in the phylogenetic
context while also estimating the off-diagonals of $\textbf{A}$, so in our model
we assume that they are zero. This is equivalent to the assumption made in 
Pagel's [-@Pagel1994] method that multiple traits cannot transition together 
instantaneously.

Following @Driver2017, the solution to \autoref{eq:ou-evo-model} for any time 
interval $t - t_0$ is

\begin{equation}
  \label{eq:solution}
  \begin{aligned}
    \eta(t) = e^{\textbf{A}(t-t_0)}\eta(t_0)\textbf{A}^{-1}[e^{\textbf{A}
    (t-t_0)}]\textbf{b} + \int_{t_0}^{t}e^{\textbf{A}(t-s)}\textbf{G}dW(s) \\
    \text{cov}\biggl[\int_{t_0}^{t}e^{\textbf{A}(t-s)}\textbf{G}dW(s)\biggr] =
    \text{irow}\biggl(\textbf{A}_\text{\#}^{-1}\Bigl[e^{\textbf{A}_\text{\#}
    (t-t_0)}- \textbf{I}\Bigr]\text{row}(\textbf{Q})\biggr)
  \end{aligned}
\end{equation}

where $\textbf{A}_{\text{\#}}=(\textbf{A}\otimes\textbf{I})+(\textbf{I}\otimes\textbf{A})$, 
with $\otimes$ denoting the Kronecker product. $\textbf{I}$ is an identity 
matrix, $\text{row()}$ is an operation that takes elements of a matrix row-wise
and puts them in a column vector, and $\text{irow()}$ is the inverse of the row
operation [@Driver2017]. In the OU process, $\frac{dy}{dt}=0$ when $y=\theta$ 
(Box \autoref{box:ou-model}). Equivalently, we can calculate the equilibrium 
value $\theta$ for each trait $\eta_i$ in the DPM as

\begin{equation}
  \label{eq:equilibrium}
  \theta_{\eta_{i}} = \frac{-(\sum_{j \neq i}\textbf{A}[i,j]\eta_j) + 
  \textbf{b}_i}{\textbf{A}[i,i]}
\end{equation}

Note that the vector $\theta_{\eta}$ does not equal the time asymptotic trait 
values for the system as a whole ($\textbf{b}_{\Delta\infty}$; *CITE*), which 
are determined by

\begin{equation}
\label{eq:asymptotic}
  \textbf{b}_{\Delta\infty} = \textbf{A}^{-1}\textbf{b}
\end{equation}

@@

Our strategy for mapping this model onto a phylogenetic tree is described in 
\autoref{alg:alg} and visualized in \autoref{fig:dpm-algorithm}. In short, we
iterate over segments of the tree and allow the coevolutionary process to play
out over the length of each tree segment.

\begin{algorithm}
  \caption{Implementation of the dynamic phylogenetic model}\label{alg:alg}
  \DontPrintSemicolon
  \For{$n \in N$}{
    Divide the evolutionary history of $n$ into $S$ segments, where each 
    segment starts with a parent node and ends with a child node or tip.\;
    Calculate the length of each segment $s \in S$, i.e., duration of time 
    between the parent and child nodes.\;
    Initialize the ancestral trait values $\eta_0$.\;
    \For{$s \in S$}{
      Solve for $\eta(s)$:
      \begin{equation}
        \nonumber
        \eta(s) = e^{\textbf{A}(t(s))}\eta(s-1)\textbf{A}^{-1}[e^{\textbf{A}
        (t(s)-t(s-1))}]\textbf{b} + \int_{t(s-1)}^{t(s)}e^{\textbf{A}
        (t(s-1)-t(s))}\textbf{G}dW(t(s))
      \end{equation}
      \;
    }
  }
\end{algorithm}

```{r dpm-algorithm, echo=FALSE, fig.align='center', fig.cap='\\textit{The DPM algorithm.} Example of how our phylogenetic algorithm works (see \\autoref{alg:alg} for formal structure). The evolutionary history of each tip (A-E) is described as a time series starting at \\(t(s)\\) and comprised of a set of segments \\(S\\). The OU process runs over each segment in order, where the initial trait values for any segment \\(s\\) are the terminal values from \\(s-1\\). Each sequence is a combination of shared segments (where tips are evolving together) and unshared segments (where tips are evolving independently). For example, S(A) and S(B) start with the same trait values because they both descend directly from the parent segment S(A,B). But further evolution along S(A) and S(B) happens independently. Dashed lines indicate segment split points.', out.width='100%'}
knitr::include_graphics("figures/dpm_algorithm.png")
```

We have implemented this algorithm in a Bayesian inferential framework using the 
Stan programming language [@Carpenter2017] (see Supplementary Materials for
example Stan code). Stan employs the computationally efficient Hamiltonian Monte
Carlo MCMC algorithm to sample from the posterior distribution. Implementing the 
model in Stan also allows for a greater degree of model flexibility. Users can
fit the model to any number of traits following any response distribution,
lifting the limitation of Pagel's [-@Pagel1999] method to binary traits and
eschewing the need for false dichotimization. Stan also allows for additional 
features, such as the incorporation of regularization through prior 
distributions, the inclusion of measurement error on traits, and handling 
missing data.

## The `coevolve` R package

To aid researchers in applying the DPM to their own data, we have developed the 
`coevolve` R package (https://github.com/ScottClaessens/coevolve). This 
fully documented package provides a user-friendly interface for applying DPMs, 
with simple functions for generating DPM Stan code and data lists, fitting DPMs
to data using the `cmdstanr` package [@R_cmdstanr], and post-processing and 
plotting model results.

The development version of the `coevolve` package can be installed and loaded
using the following R code:

```{r eval=FALSE}
#install.packages("devtools")
library(devtools)
install_github("ScottClaessens/coevolve")
library(coevolve)
```

Currently, the `coevolve` package supports binary, ordinal, count, continuous, 
and positive real traits. The package can also handle missing trait data, 
repeated observations per taxa, and model comparison using approximate 
leave-one-out cross-validation. We plan to implement further response 
distributions and latent variables in future versions of the package, but for 
now users are encouraged manually adapt the Stan code produced by the package to
add these features.

@@

In order to demonstrate that our formal model is able to accurately recover true
coevolutionary dynamics, we apply the `coevolve` package to a simple synthetic
example.

Consider an ancestral species with two continuous traits. As an illustration, 
let's assume that we are studying the highly diverse cichlid fish of Lake 
Tanganyika, and the two continuous traits are female promiscuity and sperm size
(\autoref{fig:synthetic}a). Our hypothesis is that changes in female promiscuity
will select for changes in sperm size, but not vice versa [@Fitzpatrick2009]. We
simulate the coevolution of these traits over time, allowing traits to have 
autoregressive effects (i.e., change in a trait influences itself in the 
future), coevolutionary effects (i.e., change in a trait influences the other 
trait in the future), and stochastic drift. In this particular simulation, we 
specify that changes in female promiscuity influence future changes in sperm 
size, but not vice versa.

The data generating model is as follows:

\begin{equation}
  \begin{aligned}
    \text{spermsize}_t &= (0.90 \times \text{spermsize}_{t-1}) + 
    (0.85 \times \text{promiscuity}_{t-1}) + \epsilon_{\text{spermsize}} \\
    \text{promiscuity}_t &= (0.90 \times \text{promiscuity}_{t-1}) + 
    \epsilon_{\text{promiscuity}} \\
    \epsilon_{\text{spermsize}}, \epsilon_{\text{promiscuity}} &\sim 
    \mathcal{N}(0,0.05)
  \end{aligned}
\end{equation}

We allow this coevolutionary process to play out over discrete timesteps. In 
each timestep, there is a small probability ($p = 0.05$) of a speciation event, 
where the parent species splits into two child species. Coevolutionary dynamics
then continue independently on each separate phylogenetic branch. We continue 
this process of evolution and speciation until we have a sufficient sample size
of cichlid species.
